---
title: "Getting Started with lstparsR"
output:
  rmarkdown::html_vignette:
    css: themakr-vignette.css
vignette: >
  %\VignetteIndexEntry{Getting Started with lstparsR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(lstparsR)
```

## Overview

**lstparsR** reads NONMEM `.lst` listing files and extracts parameter
estimates into tidy tibbles. It is designed for population pharmacokinetic
and pharmacodynamic (PK/PD) workflows where many model runs need to be
parsed programmatically.

Key design choices:

- All output is returned as tibbles with consistent column names.
- Missing quantities (e.g., standard errors when the covariance step is
  absent) are `NA`, never errors.
- Both VECTOR blocks (THETA) and lower-triangular MATRIX blocks
  (OMEGA, SIGMA) are handled, including multi-line parameter sets.

## Reading a Listing File

Use `read_lst_file()` to load a `.lst` file into an S3 object of class
`"lst"`:

```{r read}
path <- system.file("testdata", "full_cov.lst", package = "lstparsR")
lst  <- read_lst_file(path)
lst
summary(lst)
```

The `lst` object is simply a character vector (one element per line) with
class `"lst"` attached. All `fetch_*` functions require this class.

## Extracting THETA (Fixed Effects)

```{r thetas}
fetch_thetas(lst)
```

Each row corresponds to one THETA parameter. The `se` and `rse` columns
contain the standard error and relative standard error (as a percentage),
respectively.

Use the `digits` argument to round RSE values:

```{r thetas-rounded}
fetch_thetas(lst, digits = 1)
```

## Extracting OMEGA (Random Effects)

`fetch_etas()` extracts the diagonal of the OMEGA covariance matrix,
plus ETA shrinkage when reported:

```{r etas}
fetch_etas(lst)
```

Shrinkage values can be rounded separately:

```{r etas-rounded}
fetch_etas(lst, digits = 1, shk_digits = 1)
```

## Extracting SIGMA (Residual Error)

```{r sigmas}
fetch_sigmas(lst)
```

## Objective Function Value

```{r ofv}
fetch_ofv(lst)
```

For failed runs, `fetch_ofv()` falls back to the workflow footer
(`OFV = ...`) if the standard `#OBJV:` line is absent.

## Condition Number

The condition number is computed as the ratio of the largest to smallest
eigenvalue of the correlation matrix of parameter estimates. Values above
1000 suggest numerical instability:

```{r condn}
fetch_condn(lst)
```

## All at Once

`fetch_all()` runs every parser in a single call. Individual parser
failures are caught and returned as `NULL` with a warning, so a single
problematic section does not abort the entire extraction:

```{r fetch-all}
result <- fetch_all(lst)
names(result)
result$thetas
result$ofv
result$condn
```

## Runs Without a Covariance Step

When NONMEM did not run or complete a covariance step, SE, RSE, shrinkage,
and condition number are returned as `NA`:

```{r no-cov}
path2 <- system.file("testdata", "theta_no_cov.lst", package = "lstparsR")
lst2  <- read_lst_file(path2)
fetch_thetas(lst2)
fetch_condn(lst2)
```

## Batch Processing

A common pattern is to parse many `.lst` files at once:

```{r batch, eval = FALSE}
library(purrr)

lst_files <- list.files("models/", pattern = "\\.lst$", full.names = TRUE)

results <- map(lst_files, \(f) {
  lst <- read_lst_file(f)
  fetch_all(lst)
})

# Combine all THETA tables
all_thetas <- map_dfr(results, "thetas", .id = "run")
```

## Supported Estimation Methods

lstparsR detects and supports the following NONMEM estimation methods:

- First Order (FO)
- First Order Conditional Estimation (FOCE)
- First Order Conditional Estimation with Interaction (FOCE-I)
- Stochastic Approximation Expectation-Maximization (SAEM)
- Importance Sampling (IMP)
- Importance Sampling Assisted by Mode A Posteriori (IMPMAP)
- Markov Chain Monte Carlo Bayesian Analysis

## Interactive Exploration

For interactive use, launch the built-in Shiny application:

```{r shiny, eval = FALSE}
lstparsR::run_app()
```

This opens a browser-based interface where you can upload `.lst` files,
view parsed results in tables, and download them as CSV or RDS.
