---
title: "lstparseR: Tidy Parsing of NONMEM Listing Files"
author:
  - Raban Heller^[Charit\'{e} -- Universit\"{a}tsmedizin Berlin, raban.heller@charite.de]
  - Simeon R\"{u}desheim^[Saarland University, simeon.ruedesheim@uni-saarland.de]
  - Dominik Selzer^[Saarland University, dominik.selzer@uni-saarland.de]
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: pdflatex
    keep_tex: false
    fig_caption: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{hyperref}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{lstparseR v0.1.0}
  - \fancyhead[R]{\thepage}
  - \fancyfoot[C]{}
  - \renewcommand{\headrulewidth}{0.4pt}
vignette: >
  %\VignetteIndexEntry{lstparseR: Tidy Parsing of NONMEM Listing Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  out.width = "100%"
)
library(lstparseR)
```

\newpage

# Introduction

Population pharmacokinetic and pharmacodynamic (PK/PD) modelling with
NONMEM produces listing files (`.lst`) that contain parameter estimates,
standard errors, objective function values, eigenvalues, and shrinkage
statistics in a structured but complex text format. Extracting these
quantities for downstream analysis --- model comparison tables, forest
plots, covariate analyses --- typically requires manual copy-paste or
ad-hoc scripts that are fragile and error-prone.

**lstparseR** provides a clean, validated interface for reading `.lst`
files and returning tidy data frames. It is designed with two goals:

1. **Robustness.** All parsers return `NA` for unavailable quantities
   rather than raising errors. This makes the package safe for batch
   workflows where some NONMEM runs may have failed or been terminated
   early.

2. **Tidiness.** All output is returned as tibbles with consistent,
   lowercase column names (`parameter`, `estimate`, `se`, `rse`,
   `shrinkage`) ready for binding across runs with `dplyr` or `purrr`.

The package supports all major NONMEM estimation methods: FOCE-I, FOCE,
FO, SAEM, IMP, IMPMAP, and Bayesian analysis.

# Installation

Install the development version from GitHub:

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("Clinical-Pharmacy-Saarland-University/lstparseR")
```

# Quick Start

```{r quickstart}
library(lstparseR)

path <- system.file("testdata", "full_cov.lst", package = "lstparseR")
lst  <- read_lst_file(path)
lst
```

```{r quickstart-all}
result <- fetch_all(lst)
names(result)
```

# Reading Listing Files

`read_lst_file()` reads a `.lst` file into an S3 object of class `"lst"`.
This is simply a character vector (one element per line) with the class
attribute attached. All `fetch_*` functions require this class as input.

```{r read}
lst <- read_lst_file(
  system.file("testdata", "full_cov.lst", package = "lstparseR")
)
summary(lst)
```

# Extracting Parameters

## THETA --- Fixed Effects

`fetch_thetas()` extracts the THETA parameter estimates from the
*FINAL PARAMETER ESTIMATE* section. When the covariance step was
completed, standard errors (SE) and relative standard errors (RSE, in
percent) are also reported.

```{r thetas}
fetch_thetas(lst)
```

The `digits` argument rounds RSE values:

```{r thetas-round}
fetch_thetas(lst, digits = 1)
```

## OMEGA --- Random Effects

`fetch_etas()` extracts the diagonal of the OMEGA covariance matrix
(variance of inter-individual random effects). ETA shrinkage is included
when reported by NONMEM.

```{r etas}
fetch_etas(lst)
```

Both RSE and shrinkage rounding can be controlled independently:

```{r etas-round}
fetch_etas(lst, digits = 1, shk_digits = 1)
```

## SIGMA --- Residual Error

`fetch_sigmas()` extracts the diagonal of the SIGMA covariance matrix
(variance of residual error terms).

```{r sigmas}
fetch_sigmas(lst)
```

## Objective Function Value

`fetch_ofv()` extracts the OFV from the `#OBJV:` line. For failed or
early-terminated runs, it falls back to the workflow footer
(`OFV = ...`) if present.

```{r ofv}
fetch_ofv(lst)
fetch_ofv(lst, digits = 2)
```

## Condition Number

`fetch_condn()` computes the condition number of the correlation matrix
of parameter estimates as the ratio of the largest to the smallest
eigenvalue. Condition numbers above 1000 suggest numerical instability
in the estimation.

```{r condn}
fetch_condn(lst)
fetch_condn(lst, digits = 1)
```

## All Parameters at Once

`fetch_all()` is a convenience wrapper that calls every parser and returns
a named list. Individual failures are caught and returned as `NULL` with
a warning, so one problematic section does not abort the entire call.

```{r fetchall}
result <- fetch_all(lst)
result$thetas
result$ofv
result$condn
```

# Handling Runs Without a Covariance Step

When NONMEM did not complete the covariance step, SE, RSE, shrinkage,
and condition number are returned as `NA`. The parameter estimates and
OFV are still available.

```{r nocov}
path2 <- system.file("testdata", "theta_no_cov.lst", package = "lstparseR")
lst2  <- read_lst_file(path2)
summary(lst2)
```

```{r nocov-thetas}
fetch_thetas(lst2)
```

```{r nocov-condn}
fetch_condn(lst2)
```

# Batch Processing

A common pattern is to parse many `.lst` files from a model development
workflow:

```{r batch, eval = FALSE}
library(purrr)

files <- list.files("models/", pattern = "\\.lst$", full.names = TRUE)

results <- map(files, \(f) {
  lst <- read_lst_file(f)
  fetch_all(lst)
})

# Combine all THETA tables with run identifier
all_thetas <- map_dfr(results, "thetas", .id = "run")

# Extract OFV for model comparison
ofvs <- map_dbl(results, "ofv")
```

Since `fetch_all()` uses `purrr::safely()` internally, failed runs
produce `NULL` elements rather than stopping the pipeline.

# Function Reference

\begin{longtable}{lp{9cm}}
\toprule
\textbf{Function} & \textbf{Description} \\
\midrule
\texttt{read\_lst\_file()} & Read a \texttt{.lst} file into an \texttt{lst} object \\
\texttt{fetch\_thetas()} & THETA estimates with SE and RSE \\
\texttt{fetch\_etas()} & OMEGA diagonal with SE, RSE, and ETA shrinkage \\
\texttt{fetch\_sigmas()} & SIGMA diagonal with SE and RSE \\
\texttt{fetch\_ofv()} & Objective function value (with footer fallback) \\
\texttt{fetch\_condn()} & Condition number from eigenvalues \\
\texttt{fetch\_all()} & Run all parsers; return named list \\
\texttt{run\_app()} & Launch interactive Shiny application \\
\bottomrule
\end{longtable}

# Supported Estimation Methods

lstparseR automatically detects the estimation method from the `.lst`
file header. The following methods are supported:

\begin{itemize}
  \item First Order (FO)
  \item First Order Conditional Estimation (FOCE)
  \item First Order Conditional Estimation with Interaction (FOCE-I)
  \item Stochastic Approximation Expectation-Maximization (SAEM)
  \item Importance Sampling (IMP)
  \item Importance Sampling Assisted by Mode A Posteriori (IMPMAP)
  \item Markov Chain Monte Carlo Bayesian Analysis
\end{itemize}

# Interactive Shiny Application

For interactive exploration, lstparseR includes a Shiny application that
can be launched with:

```{r shiny, eval = FALSE}
lstparseR::run_app()
```

The application provides:

- Multi-file upload with instant parsing
- Overview dashboard with parse status summary
- Sortable tables for THETA, OMEGA, and SIGMA parameters
- OFV and condition number summary
- Download buttons for CSV and RDS export

# Session Info

```{r sessioninfo}
sessionInfo()
```
